Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-30 paper.tex

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Tom's Institute}
\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
University of Oxford, UK}
\affil[6]{Wellcome Sanger Institute}
\affil[7]{Genomics England}
\affil[8]{School of Computer Science, McGill University, Montreal, QC, Canada}

@@

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.
Recent additions using Zarr include the application of deep learning
models to genomic sequence data~\citep{klie2023predictive} 
and a browser for genetic variation data~\citep{konig2023divbrowse}.
Zarr has in fact been in use in genomics for many years, and was originally 
developed to store data from the \textit{Anopheles gambiae} 
1000 Genomes Project~\citep{anopheles2017genetic}.
A successor project, the MalariaGEN Vector Observatory,
currently uses Zarr to store data from whole-genome sequencing 
of 23,000 \textit{Anopheles} mosquitoes from 31 African 
countries~\cite{mvgo2024}.
The data is hosted in Google Cloud Storage and can be analysed interactively
using free cloud computing services like Google Colab,
democratising the use of data by scientists in malaria-endemic countries
where access to local computing infrastructure and sufficient network 
bandwidth to download large datasets may be limited.

@@

An important aspect of VCF Zarr is that field dimensions are 
global and fixed, and defined as the maximum across all rows.
Continuing the example above, the third dimension of the array 
is the maximum number of alternate alleles across \emph{all} 
variants. For variants at which there are less than the maximum
number of alternative alleles, the third dimension of the 
\texttt{call\_AD} array
is padded with a sentinel value (-2 for integers and a specific
non-signalling NaN for floats). While this is not a problem 
in practise for datasets in which all four bases are observed, 
it is a substantial issue for
fields that have a quadratic dependency on the number of alleles
(Number=G) such as PL. Such fields are already known to cause 
significant problems, and the ``local alleles'' proposal 
provides an elegant solution~\cite{poterba2024scalable}.
As this approach is on a likely path to 
standardisation~\cite{danecek2021twelve}, we would aim to 
include support in later versions of VCF Zarr.

@@

The \texttt{shuffle} parameter in the Blosc meta-compressor~\cite{alted2010modern}
can result in substantially better compression, albeit at the 
cost of somewhat slower decoding (see Fig~\ref{fig-whole-matrix-decode}).
Fig~\ref{fig-compression-shuffle} shows the effect of 
bit shuffle (grouping together bits at the same position
across bytes before compression), and byte shuffle (grouping together 
bytes at the sample position across words before compression)
on compression ratio. Bit shuffle provides a significant
improvement in compression for the call\_genotype field  because 
the vast majority of genotype calls will be 0 or 1, and therefore 
bits 1 to 7 will be 0. Thus, grouping these bits together will 
lead to significantly better compression. This strategy also 
works well when compressing boolean fields stored as 
8 bit integers, where the top 7 bits are always 0. In practise,
boolean fields stored in this way have very similar compression
to using a bit-packing pre-compression filter (data not shown).
Although byte shuffle leads to somewhat better compression for call\_AD
and call\_DP, it gives substantially worse compression on call\_AB
than no shuffling. 
The default in \texttt{vcf2zarr}
is therefore to use bit shuffle for \texttt{call\_genotype}
and all boolean fields, and to not use 
byte shuffling on any field. These defaults can be easily
overruled, however, by outputting and modifying a JSON formatted 
storage schema before encoding to Zarr.

>>>>

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Tom White Consulting Ltd.}
\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
University of Oxford, UK}
\affil[6]{Wellcome Sanger Institute}
\affil[7]{Genomics England}
\affil[8]{School of Computer Science, McGill University, Montreal, QC, Canada}

@@

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.
Recent additions using Zarr include the application of deep learning
models to genomic sequence data~\citep{klie2023predictive}, storage and manipulation of large-scale Linkage-Disequilibrium (LD) matrices \cite{ZABAD2023741},  
and a browser for genetic variation data~\citep{konig2023divbrowse}.
Zarr has in fact been in use in genomics for many years, and was originally 
developed to store data from the \textit{Anopheles gambiae} 
1000 Genomes Project~\citep{anopheles2017genetic}.
A successor project, the MalariaGEN Vector Observatory,
currently uses Zarr to store data from whole-genome sequencing 
of 23,000 \textit{Anopheles} mosquitoes from 31 African 
countries~\cite{mvgo2024}.
The data is hosted in Google Cloud Storage and can be analysed interactively
using free cloud computing services like Google Colab,
democratising the use of data by scientists in malaria-endemic countries
where access to local computing infrastructure and sufficient network 
bandwidth to download large datasets may be limited.

@@

An important aspect of VCF Zarr is that field dimensions are 
global and fixed, and defined as the maximum across all rows.
Continuing the example above, the third dimension of the array 
is the maximum number of alternate alleles across \emph{all} 
variants. For variants at which there are less than the maximum
number of alternative alleles, the third dimension of the 
\texttt{call\_AD} array
is padded with a sentinel value (-2 for integers and a specific
non-signalling NaN for floats). While this is not a problem 
in practice for datasets in which all four bases are observed, 
it is a substantial issue for
fields that have a quadratic dependency on the number of alleles
(Number=G) such as PL. Such fields are already known to cause 
significant problems, and the ``local alleles'' proposal 
provides an elegant solution~\cite{poterba2024scalable}.
As this approach is on a likely path to 
standardisation~\cite{danecek2021twelve}, we would aim to 
include support in later versions of VCF Zarr.

@@

The \texttt{shuffle} parameter in the Blosc meta-compressor~\cite{alted2010modern}
can result in substantially better compression, albeit at the 
cost of somewhat slower decoding (see Fig~\ref{fig-whole-matrix-decode}).
Fig~\ref{fig-compression-shuffle} shows the effect of 
bit shuffle (grouping together bits at the same position
across bytes before compression), and byte shuffle (grouping together 
bytes at the sample position across words before compression)
on compression ratio. Bit shuffle provides a significant
improvement in compression for the call\_genotype field  because 
the vast majority of genotype calls will be 0 or 1, and therefore 
bits 1 to 7 will be 0. Thus, grouping these bits together will 
lead to significantly better compression. This strategy also 
works well when compressing boolean fields stored as 
8 bit integers, where the top 7 bits are always 0. In practice,
boolean fields stored in this way have very similar compression
to using a bit-packing pre-compression filter (data not shown).
Although byte shuffle leads to somewhat better compression for call\_AD
and call\_DP, it gives substantially worse compression on call\_AB
than no shuffling. 
The default in \texttt{vcf2zarr}
is therefore to use bit shuffle for \texttt{call\_genotype}
and all boolean fields, and to not use 
byte shuffling on any field. These defaults can be easily
overruled, however, by outputting and modifying a JSON formatted 
storage schema before encoding to Zarr.
